---
title: "Live code:"
date: "February 14, 2023"
description: "Intro to base R"
editor: visual
categories: "Live code"
format:
  html: 
    code-fold: false
---

This lab is intended to re-familiarize yourself with R and RStudio, as well as begin practicing to code in base R. You will need the `tidyverse` package.

```{r}
#| warning: false
#| message: false
library(tidyverse)
```

## Vectors

In R, a vector is a data structure that holds or stores elements of the same type. Type may be numeric, integer, character, boolean, etc.

### The `c()` function

Generally, we create vectors using the `c()` function and then save the vector into a variable. In the code below, I create a vector of three values (10, 11, and 12), and save the results into `v`:

```{r}
v <- c(10, 11, 12)
```

### The `:` operator

Now, sometimes it's really useful to create a vector of consecutive numbers, for example, the values 1 through 10. Rather than type out every number explicitly and wrap it in `c()` , I can use the `:` operator, which looks like `a:b` where `a` and `b` are *integers* of your choosing. If `a < b`, `R` will then create a vector of numbers `a, a+1, a+2,…, b-1, b` .

```{r}
x <- 1:10
x
```

What do you think happens if `a > b`? Try the following code for yourself:

```{r}
y <- 10:1
```

### The `rep()` function

One function that I personally use a lot to create vectors is the `rep(a,b)` function, which takes in two argument. The first is the value `a` you wish to `rep`eat, and the second argument is the number of times `b` you'd like to `rep`eat it. How would we create a vector of 20 0's? Think about it, and check:

```{r}
#| eval: false
#| code-fold: true
rep(0, 20)
```

## Matrices

Matrices are the 2D extension of the one-dimensional vectors. When a matrix has `n` rows and `p` columns, we denote its dimensions as `n x p` or "n by p". We create matrices using the `matrix()` function. Because of the multiple dimensions, we need to specify the number of rows and the number of columns:

```{r}
matrix(NA, nrow = 2, ncol = 3)
```

This code above creates a `2 x 3` matrix of `NA` values. The first argument takes in the elements you want to fill the matrix with. This can either be a single value, or a single vector of values.

## Data frames

We will create a data frame called `my_df` here, which holds the two vectors we created before.

```{r}
my_df <- data.frame(xvar = x, yvar = y)
```

Now, if I wanted to only take the variable `xvar` from `my_df`, how would I do so using `dplyr` functions? Take a second to think about it, then check:

```{r}
#| code-fold: true 
#| eval: false

my_df %>%
  select(xvar)

```

We will now use base R to access that `xvar` variable by using `$` notation: `<df>$<var_name>` . If you do this yourself, you should notice that immediately after typing the `$` , a menu pops up with all the variables contained in the data frame.

```{r}
#| eval: false
my_df$xvar
```

Now, do you notice the difference between the two outputs?

```{r}
my_df %>%
  select(xvar)

my_df$xvar
```

## Indexing

One of the most useful tools we will use is *indexing* and index notation.

An index is essentially a numerical representation of an item's position in a list or vector. It is typically an integer, starting from either 0 or 1 depending on the programming language. In R, our index positions always start at 1!

For example, in the word "learning", the l is at index 1. Similarly, in our vector `v` of the numbers $(10, 11, 12)$, the value at index 1 is 10. We can confirm this with code:

```{r indexing-1}
v[1]
```

Notice that we access the item held in index 1 using the **square bracket notation** `[ ]`

Now, I can also **replace** or **modify** an element at a given index. I will still access the location using `[ ]`, but now I will store/save the new value:

```{r indexing-2}
v[1] <- 13
v
```

We can also modify multiple elements at once by passing in a vector of indices to modify, as well as a vector of new values:

```{r}
v[2:3] <- c(14, 15)
```

What does `v` look like now?

We can also use indices to refer to elements or entire *rows* and *columns* of data frames! Unlike vectors, data frame are **two-dimensional**, i.e. there are both rows and columns. Thus, our index notation will need to adapt to accommodate this feature. We will still use `[ ]` notation, but now commas will be introduced:

```{r}
my_df[1,2]
```

Based on `my_df`, what do you think the `[1,2]` means?

Now, we already saw how to access a column of a data frame using the `$` notation, but we can also use index notation. To access the first column, we would type:

```{r}
#| eval: false
my_df[,1]
```

The `1` after the comma tells `R` that we want to focus on column `1`.

As I do not type anything before the comma, `R` reads this as: "since you did not want a specific row, you must want all the rows".

How do you think we would access the third column? How about both the first and second row?

## Functions

There are a lot of simple functions in `R` that we will rely on. We already saw `c()` and `rep()`. Most of the functions we will use take in a vector or matrix of numeric values, and return either a single number or vector in return.

The function `mean()` takes in a vector, and returns the mean of the vector.

```{r}
mean(x)
```

The function `length()` takes in a vector, and returns the number of elements in the vector:

```{r}
length(x)
```

The functions `max()` and `min()` return what you would expect them to!

An extremely useful function we will use is the `which()` function. Unlike the previous functions, `which()` does not take in a numeric vector. Rather, it takes a vector of **boolean** values (i.e. TRUE/FALSE values). Then, it returns the indices of the TRUE values in the vector. For example, an input might be:

```{r}
y == -5
```

This is comparing each value in `y` to see if it equals `-5` (recall the double equals sign check for equality). Notice that only one value evaluates to TRUE, specifically the element in index 6. Therefore, if we *wrap* the `which()` function around that argument, we should get `6` in return:

```{r}
which(y == -5)
```

Personally, I tend to read this line of code as a question: Which element(s) of `y` are exactly equal to `-5`?

We know that `y` only holds negative values. What do you think happens if we try to evaluate the following. Try it yourself!

```{r}
#| eval: false
which(y == 0)
```

It's also entirely possible that many values in the boolean vector are true, in which case the function would return multiple indices. For example, if I want to know which values in `y` are negative, I could code:

```{r}
which(y < 0)
```

### Indexing with boolean vectors

Above, we saw how to access elements held at specific indices of interest. We can also use boolean vectors to return values. Recall our vector `v`:

```{r}
v
```

I can index `v` by indexing using `TRUE`'s for each value that I want, and `FALSE`'s otherwise.

```{r}
v[c(F, T, F)]
```

## Your turn!

Please complete the following exercises in order:

1.  Create a vector called `my_vec` that holds the values 50 through 100.
2.  Create a new vector called `less60` where an element is `TRUE` if the corresponding element in `my_vec` is less than `60`, and `FALSE` otherwise.
3.  Confirm that the length of your two vectors are the same.
4.  Pass `less60` into the function `sum()` function. Relate the value obtained to the elements of `less60`.
5.  Modify `my_vec` such that the value at index 10 is 100.
6.  Obtain the index of the maximum values of `my_vec` using functions described above.
7.  Now, pass `my_vec` into the `which.max()` function. Even though we haven't seen it before, based on the function name, the name of the function is intuitive. Does the result from `which.max()` differ from what you obtained in Ex. 6? How so?
8.  Create a `2 x 5` matrix of the values 1 through 10, where the first row holds the values 1-5, and the second row holds the values 6-10. *Hint*: look at the help file for `matrix`.
