---
title: "Linear Regression"
date: ""
format: 
  revealjs:
    theme: custom.scss
    transition: none
    incremental: true
editor: visual
editor_options: 
  chunk_output_type: console
draft: true
---

# Housekeeping

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = F, warning = F, message = F)

library(tidyverse)
library(broom)
library(ggfortify)
library(ISLR)
library(vegan)
plot_theme <- theme(text = element_text(size = 14))
data(mite)
data(mite.env)
mite_dat <- mite.env %>%
  add_column(abundance = mite$LRUG)
```

# Linear regression

## Linear regression

-   A simple, widely used approach in supervised learning

-   Assumes that the dependence of $Y$ on the predictors $X_{1}, \ldots, X_{p}$ is linear

```{r fig.align="center",fig.width=8, fig.height=5}
mite_dat %>%
  pivot_longer(cols = 1:2, names_to = "variable", values_to = "value") %>%
  ggplot(., aes(x = value, y = abundance)) +
  geom_point() +
  facet_wrap(~variable, scales = "free") +
  geom_smooth(method = "lm", se = F) +
  plot_theme +
  ylab("LRUG abundance")
```

## Possible questions of interest

-   Is there a relationship between the abundance of LRUG mites and substrate density or water content in the soil where they are found?

    -   If so, how are strong are these relationships?

-   Is the relationship linear?

-   How accurately can we predict the abundance of these mites?

## Simple linear regression

-   ::: important
    Simple linear regression (SLR)
    :::

    -   Assumes a linear model for a quantitative response $Y$ using a *single* predictor $X$

        $$Y = \beta_{0} + \beta_{1} X + \epsilon,$$

        where $\beta_{0}, \beta_{1}$ are unknown **coefficients** (**parameters**) and $\epsilon$ is the error

-   $\beta_{0}$ is commonly referred to as the **intercept**, and $\beta_{1}$ is the **slope**

-   For example: `abundance` = $\beta_{0}$ + $\beta_{1}$ `watercont` + $\epsilon$

## Parameter estimation

-   Assuming $n$ observations, we have data of the form $(x_{1}, y_{1}), (x_{2}, y_{2}), \ldots, (x_{n}, y_{n})$

-   An SLR model says $$y_{i} = \beta_{0} + \beta_{1}x_{i}+ \epsilon \approx \beta_{0} + \beta_{1}x_{i}, \quad \text{ for all } i = 1,\ldots, n$$

    -   In this model, $f(x_{i}) = \beta_{0} + \beta_{1} x_{i}$

    -   Notice that the relationship between $x_{i}$ and $y_{i}$ is the same for all $i$

-   In practice, $\beta_{0}$ and $\beta_{1}$ are unknown, so we must estimate them

-   **Goal**: obtain (good) estimates $\hat{\beta}_{0}$ and $\hat{\beta}_{1}$ that are as close to the true values as possible, such that $y_{i} \approx \hat{\beta}_{0} + \hat{\beta}_{1} x_{i}$
