---
title: "Welcome to statistical learning!"
format: 
  revealjs:
    theme: custom.scss
    transition: none
editor: visual
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
opts_chunk$set(fig.height = 2.65, dpi = 300) 
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
	fig.align = "center",
	fig.height = 3.75,
	fig.width = 6.25,
	message = FALSE,
	warning = FALSE,
	echo = TRUE
)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```

## Necessary background

-   I assume you have taken Math 118 prior to this course, and are comfortable with `tidyverse` and RStudio. There is a large emphasis on computing.

-   We will learn how to code in base `R`, and by the end of the course you should feel comfortable switching between base `R` and `tidyverse`.

-   We will focus more on applications and developing intuition. The goal is to beging developing a toolbox of methods that you may use in future analyses.

# Your Turn!

## Create a GitHub account

::: instructions
Go to https://github.com/, and create an account (unless you already have one). After you create your account, click [here](https://forms.gle/W8Aw1odtjgL41S2u6) and enter your GitHub username.
:::

Tips for creating a username from [Happy Git with R](http://happygitwithr.com/github-acct.html#username-advice).

-   Incorporate your actual name!
-   Reuse your username from other contexts if you can.
-   Shorter is better than longer; be as unique as possible in as few characters as possible.
-   Avoid words laden with special meaning in programming, like `NA`.

# Course details

## Class Meetings

::: incremental
-   Lecture
    -   Focus on concepts behind statistical learning techniques
    -   Interactive lecture that includes examples and hands-on exercises
    -   Bring fully-charged laptop to every lecture
        -   Please let me know as soon as possible if you do not have access to a laptop
-   Lab
    -   Typically occurs on Fridays
    -   Focus on computing using functions provided in `R` packages
    -   Apply concepts from lecture to case study scenarios
-   Implementation
    -   Some days will be focused on implementing (i.e. coding by hand) methods discussed in lecture
    -   Group work
:::

## Assignments

-   Lab assignments
-   Implementation deliverables
-   Celebrations of knowledge (i.e. mini projects or larger coding assignments)

## Major assessments

::: incremental
-   One midterm with two components:
    -   Computational component (take-home)
    -   Oral component
-   Final project
    -   Groups of 3-4 students (tentatively)
    -   Presentations during last two days of class
    -   NO sit-down final
:::

## Important dates

-   Friday, 3/31: take-home midterm

-   Monday, 4/3: oral midterm

-   ::: important
    Friday, 4/14: Spring symposium (no class)
    :::

-   Monday, 4/17: final day to drop classes :(

-   ::: important
    Friday, 5/13 and Monday, 5/15: project presentations
    :::

## Code

```{r}
#| eval: false
library(tidyverse)
```
