---
title: "Welcome to statistical learning!"
format: 
  revealjs:
    theme: custom.scss
    transition: none
editor: visual
---

# Introductions

## About me

-   PhD in Statistical Science from Duke University, BA in Mathematics and Computer Science from Swarthmore College

-   Research interests: Bayesian hierarchical models for ecological applications

    -   Developing models for single species or community abundances

-   Office: Warner 214

    -   If my door is open, come on in! Also feel free to e-mail me.

    -   Office hours: TBD

-   Current hobbies: running, mushroom foraging

-   Aspirational hobbies/skills: fly fishing, driving stick shift

## About you

-   Introduce yourself using any or all of the following (the first is mandatory):

    -   Name
    -   Year
    -   Major/minor
    -   Hobbies
    -   How do you take your coffee?

# Course details

## About the course

- Course website: https://math218-spring2023.github.io/ (please bookmark!)

- Learn various models for regression and classification tasks (more on this next lecture)

  - Linear and logistic regression, KNN, decision trees + variants, K-means and hierarchical clustering
  

## Necessary background

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
opts_chunk$set(fig.height = 2.65, dpi = 300) 
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(
	fig.align = "center",
	fig.height = 3.75,
	fig.width = 6.25,
	message = FALSE,
	warning = FALSE,
	echo = TRUE
)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```

-   I assume you have taken Math 118 prior to this course, and are comfortable with `tidyverse` and RStudio. There is a large emphasis on computing.

    -   I also assume you are comfortable with knitting. In this course, I ask that you knit to PDF.

-   We will learn how to code in base `R`, and by the end of the course you should feel comfortable switching between base `R` and `tidyverse`.

-   We will focus more on applications and developing intuition. The goal is to begin developing a toolbox of methods that you may use in future analyses.

## Class Meetings

::: incremental
-   Lecture
    -   Focus on concepts behind statistical learning techniques
    -   Interactive lecture that includes examples and hands-on exercises
    -   Bring fully-charged laptop to every lecture
        -   Please let me know if you do not have access to a laptop
-   Lab
    -   Typically occurs on Fridays
    -   Focus on computing using functions provided in `R` packages
    -   Apply concepts from lecture to case study scenarios
-   Implementation
    -   Some days will be focused on implementing (i.e. coding by hand) methods discussed in lecture
    -   Complete in small groups
:::

## Assignments

1.  Lab assignments
2.  Implementation deliverables
3.  Celebrations of knowledge (i.e. mini projects or larger coding assignments)

## Major assessments

::: incremental
-   One midterm with two components:
    -   Computational component (take-home)
    -   Oral component
-   Final project
    -   Groups of 3-4 students (tentatively)
    -   Presentations during last two days of class
    -   NO sit-down final
:::

## Important dates

-   Friday, 3/31: take-home midterm

-   Monday, 4/3: oral midterm

-   ::: important
    Friday, 4/14: Spring symposium (no class)
    :::

-   Monday, 4/17: final day to drop classes :(

-   ::: important
    Friday, 5/13 and Monday, 5/15: project presentations
    :::

# Your Turn!

## Create a GitHub account

::: instructions
Go to https://github.com/, and create an account (unless you already have one). After you create your account, click [here](https://forms.gle/W8Aw1odtjgL41S2u6) and enter your GitHub username.
:::

Tips for creating a username from [Happy Git with R](http://happygitwithr.com/github-acct.html#username-advice).

-   Incorporate your actual name!
-   Reuse your username from other contexts if you can.
-   Shorter is better than longer; be as unique as possible in as few characters as possible.
-   Avoid words laden with special meaning in programming, like `NA`.

## Playing with base R

-   Create a folder on your desktop called Math218
-   Open RStudio and create a new Rmarkdown document.
    -   We will work through some coding exercises. The associated code can be found in "Live Code 01"

# TO DO

<!-- - Sign up for brief one-on-one meeting with me so I can get to know you -->

- Enter your GitHub username into the Google form

- Try the practice exercises involved with "Live code 01"

